---
title: "Trade in food, feed and drink"
always_allow_html: true
output:
  html_document:
    toc: yes
    toc_float: yes
params:
  ffddata: ffd202103.csv
  ffdcountrydata: ffdcountry202103.csv
  nrdata: nonresponse202103.csv
  date: '2021-03-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	# results = "hide",
	fig.width = 9,
	fig.height = 7
)

library("here")
library("dplyr")
library("ggplot2")
library("purrr")
library("tidyr")
library("patchwork")
library("stringr")
library("readr")
library("lubridate")
library("scales")
library("reactable")

source(here("R", "money.R"))

# chart chunks
chart_palette <- "Set2"

theme <-   theme_minimal() +
           theme(legend.title = element_blank(),
                 panel.grid.minor = element_blank())


```

```{r read_raw_data}

# Read in datasets

# FFD data
# convert date text to date
# tidy flow descriptions
# convert values from £000s to actual

ffd <- read.csv(here("data", "raw", params$ffddata),
                col.names = c("year", "month", "flow", "divcode", "divdesc", "product", "value", "tonnes")) %>% 
  mutate(date = as.Date(paste(year, month, "1", sep = "-"), format = "%Y-%m-%d"),
         flow = case_when(flow == "Total Exports" ~ "Exports",
                          flow == "Total Imports" ~ "Imports"),
         value = value * 1000) 
  # filter(divcode != 22)


ffdcountry <- read.csv(here("data", "raw", params$ffdcountrydata),
                       col.names = c("year", "month", "flow", "eunoneu", "country", "divcode", "divdesc", "value", "tonnes")) %>% 
    mutate(date = as.Date(paste(year, month, "1", sep = "-"), format = "%Y-%m-%d"), 
           flow = case_when(flow == "Total Exports" ~ "Exports",
                          flow == "Total Imports" ~ "Imports"),
           value = value * 1000)

# Non response data
# https://www.uktradeinfo.com/trade-data/ots-custom-table/?id=cdee0450-14b3-4e26-bed5-16e2e23b105a
# 
# add EU flag
# tidy divcodes to numeric and filter out div 0 (could do this and get new url)

nonresponse <- read_csv(here("data", "raw", params$nrdata),
                        col_names = c("sitc1",
                                      "sitc2",
                                      "sitc3",
                                      "sitc4",
                                      "sitc5",
                                      "flow",
                                      "year",
                                      "month",
                                      "include",
                                      "value_sterling"),
                        skip = 1) %>% 
  select(!(sitc3:sitc5)) %>% 
  mutate(date = ymd(paste(year, month, "1", sep = "-")),
         flow = case_when(flow == "EU - Exports" ~ "Exports",
                          flow == "EU - Imports" ~ "Imports"),
         eunoneu = "EU",
         divcode = case_when(str_starts(sitc2, "00") ~ 0,
                             str_starts(sitc2, "01") ~ 1,
                             str_starts(sitc2, "02") ~ 2,
                             str_starts(sitc2, "03") ~ 3,
                             str_starts(sitc2, "04") ~ 4,
                             str_starts(sitc2, "05") ~ 5,
                             str_starts(sitc2, "06") ~ 6,
                             str_starts(sitc2, "07") ~ 7,
                             str_starts(sitc2, "08") ~ 8,
                             str_starts(sitc2, "09") ~ 9,
                             str_starts(sitc2, "11") ~ 11,
                             str_starts(sitc2, "22") ~ 22,
                             str_starts(sitc2, "41") ~ 41,
                             str_starts(sitc2, "42") ~ 42,
                             str_starts(sitc2, "43") ~ 43),
         valuenr = value_sterling) %>% 
         select(date, eunoneu, flow, divcode, valuenr) %>% 
         filter(divcode != 0)


# Get div codesc and descriptions

divcodes <- read.csv(here("data", "tidy", "divcodes.csv"))


```

```{r summary_value_datasets}

# FFD totals -----------------------------------------------------------------

# FFD by divcode and month
divs <- ffd %>% 
  group_by(flow, divcode, date) %>% 
  summarise(value = sum(value)) %>% 
  filter(date >= "2018-01-01") %>% 
  ungroup()

# FFD inc nonresponse by month
ffdall <- divs %>% 
     left_join(nonresponse) %>% 
      mutate(valuetot = replace_na(value + valuenr, 0)) %>%
      group_by(date, flow) %>% 
      summarise(value = sum(valuetot)) %>% 
      ungroup()

# FFD year on year %ages
yoy <- ffdall %>%
      group_by(flow) %>%
      mutate(diff = lag(value, n = 12, order_by = date),
             yoy = (value/diff)-1) %>%
      arrange(flow) %>% 
      ungroup()

# EU totals --------------------------------------------------------------------

# NR by month with EU flag
eunr <- nonresponse %>% 
        group_by(date, flow, eunoneu) %>% 
        summarise(value = sum(valuenr)) %>% 
        ungroup

# FFD inc nonresponse by EU/non EU
eu <- ffdcountry %>% 
        # left_join(nonresponse) %>% 
        select(date, flow, eunoneu, country, divcode, divdesc, value) %>% 
        group_by(date, flow, eunoneu) %>% 
        summarise(value = sum(value)) %>% 
        rbind(eunr) %>% 
        group_by(date, flow, eunoneu) %>% 
        summarise(value = sum(value)) %>% 
        ungroup()

# FFD year on year %ags by EU flag
# values and %ages are prettified 
euyoy <- eu %>% 
        group_by(eunoneu, flow) %>% 
        mutate(diff = lag(value, n = 12),
               yoy = percent((value/diff) - 1, accuracy = 0.1),
               amount = money(value)) %>% 
        arrange(flow, eunoneu) %>%
        filter(date >= "2021-01-01") %>% 
        select(date, flow, eunoneu, amount, yoy) %>% 
        pivot_wider(names_from = date, 
                    values_from = c(amount, yoy)) %>% 
        ungroup()

# FFD by div totals ------------------------------------------------------------

# FFD inc nonresponse by div
ffddiv <- divs %>% 
     left_join(nonresponse) %>% 
     left_join(divcodes) %>% 
      mutate(valuetot = replace_na(value + valuenr, 0)) %>% 
      select(!c(value, valuenr, divcode)) %>%  
      group_by(flow, divdesc, date) %>% 
      summarise(value = sum(valuetot)) %>% 
      ungroup()

divyoy <- ffddiv %>%
  group_by(flow, divdesc) %>% 
      mutate(diff = lag(value, n = 12),
               yoy = (value/diff) - 1,
               amount = money(value)) %>% 
  select(date, flow, divdesc, amount, yoy) %>% 
  ungroup()
  


# EU/YoI/RoW totals ------------------------------------------------------------

# Region x div x month
euirl <- ffdcountry %>% 
  mutate(euirl = (case_when(eunoneu == "EU" & country == "Irish Republic" ~ "Ireland",
                            eunoneu == "EU" & country != "Irish Republic" ~ "EU exc RoI",
                            eunoneu == "Non EU" ~ "RoW"))) %>% 
        group_by(date, flow, eunoneu, euirl, divdesc) %>% 
        summarise(value = sum(value)) %>% 
        ungroup()


euirlyoy <- euirl %>% 
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   val = money(value),
                   yoy = percent((value/diff)-1, accuracy = 1)) %>% 
            filter(date >= "2021-01-01") %>%
            pivot_wider(names_from = date, values_from = c(val,yoy)) %>% 
            ungroup() %>% 
            select(!c(diff, value))

```

## `r month(params$date, label = TRUE, abbr = FALSE)` `r year(params$date)`

## Intro

-   This pack presents the first provisional estimates of trade data for **`r month(params$date, label = TRUE, abbr = FALSE)` `r year(params$date)`** and includes revised data from HMRC for previous months. Note that all data from January 2020 onwards is provisional and subject to change. Changes in data collection methods from 1 Jan 2021 mean that care should be taken in the interpretation of changes in trade before and after this date.

-   There has been an unusually high level of non-response for EU to GB agrifood imports for `r month(params$date, label = TRUE, abbr = FALSE)` `r year(params$date)`. Estimates to adjust for this non-response are included in the **value** data shown here, as published at an aggregate level by HMRC. HMRC do not publish non-response estimates for volumes.

-   Please note that monthly data can fluctuate, and small movements should be treated with caution.

-   We can't attribute changes in trade to a single cause. The combined effects of COVID-19 national and international lockdown restrictions, EU exit uncertainty and border disruptions have all been contributing factors to the differing nature of recent UK and global trade. These (and other factors) will impact over a longer time period.

## Trade by value

-   **Imports** in `r month(params$date, label = TRUE, abbr = FALSE)` were **`r yoy %>% filter(date == params$date, flow == "Imports") %>% pull(value) %>% money`**, a change of `r  yoy %>% filter(date == params$date, flow == "Imports")  %>% pull(yoy) %>% scales::percent()` from `r month(params$date, label = TRUE, abbr = FALSE)` `r year(params$date) - 1`.

-   **Exports** in `r month(params$date, label = TRUE, abbr = FALSE)` were **`r yoy %>% filter(date == params$date, flow == "Exports") %>% pull(value) %>% money`**, a change of `r  yoy %>% filter(date == params$date, flow == "Exports")  %>% pull(yoy) %>% scales::percent()` from `r month(params$date, label = TRUE, abbr = FALSE)` `r year(params$date) - 1`.

Trade has generally recovered from the downturn seen in the first two months of 2021.


**Year on year trade comparisons EU/RoW (inc NR)**

``` {r eu_yoy_table}

reactable(euyoy,
          defaultColDef = colDef(header = function(value) gsub("_", "\n ", value, fixed = TRUE)),
          columns = list(
            flow = colDef(name = "Trade flow"),
            eunoneu = colDef(name = "EU/Non EU")
          ))

```

```{r eu_charts, fig.height=4}
# 
# knitr::opts_chunk$set(
# 	fig.height = 4
# )

ffdall %>%
  filter(date >= "2019-01-01") %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line(aes(colour = flow), stat = "identity") +
  scale_y_continuous(labels = label_comma(scale = 0.000000001, prefix = "£", suffix = "bn")) +
  scale_x_date(date_labels = "%Y %b") +
  scale_colour_brewer(palette = chart_palette) + 
  theme +
  labs(x = NULL,
       y = NULL,
       title = "Value of trade in food, feed and drink (inc NR)")

# yoy %>% 
#   ggplot(aes(x = date, y = yoy)) +
#   geom_bar(aes(fill = flow), stat = "identity") +
#   facet_grid(rows = ~flow) +
#   scale_y_continuous(labels = scales::percent) +
#   scale_x_date(limits = ymd(c("2018-01-01", "2021-03-01"))) +
#   theme_minimal() +
#   labs(x = NULL,
#        y = "% change year on year",
#        title = "Percentage change in value of trade")

eu %>% 
  filter(date >= "2019-01-01") %>% 
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(colour = flow)) +
  facet_grid(rows = ~eunoneu) +
  scale_y_continuous(labels = label_comma(scale = 0.000000001, prefix = "£", suffix = "bn")) +
  scale_x_date(date_labels = "%Y %b") +
  scale_colour_brewer(palette = chart_palette) +
  theme +
  labs(x = NULL,
       y = NULL,
       title = "Value of trade, EU/Non EU (inc NR)")

# Matt suggestion
# eu %>% 
#   filter(date >= "2019-01-01") %>% 
#   mutate(x = paste(eunoneu, "-", flow)) %>% 
#   ggplot(aes(x = date, y = value)) +
#   geom_line(aes(colour = x)) +
#   scale_y_continuous(labels = label_comma(scale = 0.000000001, prefix = "£", suffix = "bn")) +
#   scale_x_date(date_labels = "%Y %b") +
#   scale_colour_brewer(palette = chart_palette) +
#   theme +
#   labs(x = NULL,
#        y = NULL,
#        title = "Value of trade, EU/Non EU (inc NR")



```

### By product

```{r product}

# ffddiv <- divs %>% 
#      left_join(nonresponse) %>% 
#      left_join(divcodes) %>% 
#       mutate(valuetot = replace_na(value + valuenr, 0)) %>% 
#       select(!c(value, valuenr, divcode)) %>%  
#       group_by(flow, divdesc, date) %>% 
#       summarise(valuetot = sum(valuetot))
# 
# divyoy <- ffddiv %>%
#   group_by(flow, divdesc) %>% 
#       mutate(diff = lag(valuetot, n = 12),
#                yoy = percent((valuetot/diff) - 1),
#                amount = money(valuetot*1000)) %>% 
#   filter(date == "2021-03-01") %>% 
#   select(flow, divdesc, amount, yoy)

# ffddiv %>%
#   filter(date >= "2020-01-01") %>% 
#   ggplot(aes(x = date, y = value)) +
#   geom_line(aes(colour = flow)) +
#   facet_grid(rows = vars(divdesc),
#              cols = vars(flow),
#              scales = "free",
#              labeller = label_wrap_gen(width = 30)) +
#   scale_y_continuous(labels = label_comma(scale = 0.000000001, prefix = "£", suffix = "bn")) +
#   scale_x_date(date_labels = "%Y \n %b") +
#   theme_minimal() +
#   theme(legend.title = element_blank(),
#         strip.text.y = element_text(angle = 0),
#         legend.position = "none") +
#   labs(x = NULL,
#        y = "value",
#        title = "Value of trade, SITC2 level")


# Value of trade time series small multiple by divcode
ffddiv %>%
  filter(date >= "2020-01-01") %>% 
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(colour = flow)) +
  facet_wrap(vars(divdesc),
             scales = "fixed",
             labeller = label_wrap_gen(width = 20)) +
scale_y_continuous(labels = label_comma(scale = 0.000001, prefix = "£", suffix = "m")) +
  scale_x_date(date_labels = "%y \n %b") +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray99", linetype = 0)
        ) +
  scale_colour_brewer(palette = chart_palette) +
  labs(x = NULL,
       y = NULL,
       title = "Value of trade, SITC2 level (inc NR)")


# Year on year change barcharts by mmonth facet by divcode, exports and imports
divyoy %>% 
  filter(date >= "2021-01-01", flow == "Exports") %>% 
  ggplot(aes(x = date, y = yoy)) +
  geom_bar(aes(fill = flow), stat = "identity", position = "dodge") +
  facet_wrap(vars(divdesc), labeller = label_wrap_gen(width = 20)) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = chart_palette) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_rect(fill = "gray97", linetype = 0),
        panel.border = element_blank()) +
  labs(x = NULL,
       y = NULL,
       title = "Year on year change in value of exports, 2021 (inc NR)")

divyoy %>% 
  filter(date >="2021-01-01", flow == "Imports") %>% 
  ggplot(aes(x = date, y = yoy)) +
  geom_bar(aes(fill = flow), stat = "identity", position = "dodge") +
  facet_wrap(vars(divdesc), labeller = label_wrap_gen(width = 20)) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = chart_palette) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_rect(fill = "gray97", linetype = 0),
        panel.border = element_blank()) +
  labs(x = NULL,
       y = NULL,
       title = "Year on year change in value of imports, 2021 (inc NR)")


```

### Table: Value of trade and year on year percentage change, SITC2 (inc NR)

``` {r divyoy_table}
divyoy %>% 
  filter(date == params$date) %>% 
  mutate(yoy = percent(yoy)) %>% 
  reactable(filterable = TRUE, compact = TRUE)

```

### Country splits

```{r countries}

euirl %>%            
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   val = money(value),
                   yoyval = value - diff,
                   yoypc = percent((value/diff)-1, accuracy = 1)) %>% 
            filter(date == "2021-03-01") %>%
            ggplot(aes(x = euirl, y = yoyval)) +
            geom_bar(aes(fill = euirl),stat = "identity") +
            facet_grid(rows = vars(divdesc), cols = vars(flow), scales = "free") +
            scale_y_continuous(labels = comma) +
            theme_minimal() +
            theme(panel.grid.minor = element_blank(),
                  legend.title = element_blank(),
                  legend.position = "bottom",
                  strip.text.y = element_text(angle = 0)) +
            labs(x = NULL,
                 y = "value",
                 title = "Year on year change (exc NR)")

 
euirl %>% 
  filter(flow == "Imports", date >= "2020-01-01") %>% 
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(colour = euirl)) + 
  facet_grid(rows = vars(divdesc),
             cols = vars(euirl),
             scales = "free",
             labeller = label_wrap_gen(width = 30)) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_labels = "%y \n %b") +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        strip.text.y = element_text(angle = 0)) +
  labs(x = NULL,
       y = "Value",
       title = "Value of imports by region (exc NR)")

euirl %>% 
  filter(flow == "Exports", date >= "2020-01-01") %>% 
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(colour = euirl)) + 
  facet_grid(rows = vars(divdesc),
             cols = vars(euirl),
             scales = "free",
             labeller = label_wrap_gen(width = 30)) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_labels = "%y \n %b") +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text.y = element_text(angle = 0)) +
  labs(x = NULL,
       y = "Value",
       title = "Value of exports by region (exc NR)")


```

```{r}

# Matt - add back RoW
euirl %>%            
            ungroup() %>% 
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   val = money(value),
                   yoyval = value - diff,
                   yoypc = percent((value/diff)-1, accuracy = 1)) %>% 
            filter(date == "2021-03-01") %>%
            ggplot(aes(x = flow, y = yoyval, group = euirl)) +
            geom_bar(aes(fill = euirl), stat = "identity") +
            facet_wrap(vars(divdesc), scales = "free", labeller = label_wrap_gen(width = 20)) +
            scale_y_continuous(labels = comma) +
            theme_minimal() +
            theme(panel.grid.minor = element_blank(),
                  legend.title = element_blank(),
                  legend.position = "bottom",
                  strip.text.y = element_text(angle = 0)) +
            labs(x = NULL,
                 y = "value",
                 title = "Year on year change in trade values, EU (exc NR)")



euirl %>%            
            ungroup() %>% 
            filter(flow == "Exports") %>% 
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   val = money(value),
                   yoyval = value - diff,
                   yoypc = percent((value/diff)-1, accuracy = 1)) %>% 
            filter(date == "2021-03-01") %>%
            ggplot(aes(x = divdesc, y = yoyval, group = euirl)) +
            geom_bar(aes(fill = euirl), stat = "identity") +
            # facet_grid(cols = vars(divdesc), scales = "free", labeller = label_wrap_gen(width = 20)) +
            scale_y_continuous(labels = comma) +
            scale_x_discrete(labels = function (x) str_wrap(x, width = 30)) +
            coord_flip() +
            theme_minimal() +
            theme(panel.grid.minor = element_blank(),
                  legend.title = element_blank(),
                  legend.position = "bottom",
                  strip.text.y = element_text(angle = 0)) +
            labs(x = NULL,
                 y = "value",
                 title = "Year on year change in export values, March 2021 (exc NR)")

euirl %>%            
            ungroup() %>% 
            filter(flow == "Imports") %>% 
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   val = money(value),
                   yoyval = value - diff,
                   yoypc = percent((value/diff)-1, accuracy = 1)) %>% 
            filter(date == "2021-03-01") %>%
            ggplot(aes(x = divdesc, y = yoyval, group = euirl)) +
            geom_bar(aes(fill = euirl), stat = "identity") +
            # facet_grid(cols = vars(divdesc), scales = "free", labeller = label_wrap_gen(width = 20)) +
            scale_y_continuous(labels = comma) +
            scale_x_discrete(labels = function (x) str_wrap(x, width = 30)) +
            coord_flip() +
            theme_minimal() +
            theme(panel.grid.minor = element_blank(),
                  legend.title = element_blank(),
                  legend.position = "bottom",
                  strip.text.y = element_text(angle = 0)) +
            labs(x = NULL,
                 y = "value",
                 title = "Year on year change in import values, March 2021 (exc NR)")



```

### Table: Values and year on year change, March 2021 (exc NR)

``` {r}
euirlyoy %>% 
reactable(filterable = TRUE, compact = TRUE)

```

### David K addition

``` {r david_kennedy}

# %ages
euirlpc <- euirl %>%
  filter(eunoneu == "EU") %>% 
  group_by(date,flow, eunoneu, divdesc) %>% 
  mutate(pc = value/sum(value))

euirlpc %>% 
  filter(flow == "Exports", date >= "2020-01-01") %>% 
  ggplot(aes(x = date, y = pc)) +
  geom_col(aes(fill = euirl), position = "stack", stat = "identity") +
  scale_x_date(date_labels = "%Y\n%b") +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = chart_palette, direction = -1) +
  facet_wrap(~divdesc, labeller = label_wrap_gen()) +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  labs(x = NULL,
       y = NULL,
       title = "Proportion of EU export values to Ireland, 2020-2021 (exc NR)")

euirlpc %>% 
    filter(flow == "Exports", date >= "2020-01-01") %>% 
    mutate(pc = percent(pc)) %>% 
    select(!c(value)) %>% 
    pivot_wider(names_from = date, values_from = pc) %>% 
    arrange(divdesc, euirl) %>% 
    reactable(filterable = TRUE, compact = TRUE)


# values
valdiffs <- euirl %>% 
            filter(date >= "2019-01-01") %>% 
            group_by(divdesc, flow, euirl) %>% 
            mutate(diff = lag(value, n = 12),
                   valuediff = value - diff) %>% 
            ungroup()

valdiffs %>% 
  filter(date >= "2021-01-01", flow == "Exports", euirl != "RoW") %>% 
  ggplot(aes(x = date, y = valuediff)) +
  geom_bar(aes(fill = euirl), stat = "identity") +
  facet_wrap(~divdesc, labeller = label_wrap_gen()) +
  scale_y_continuous(labels = label_comma(scale = 0.000001, prefix = "£", suffix = "m")) +
  scale_fill_brewer(palette = chart_palette, direction = -1) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_rect(fill = "gray97", linetype = 0),
        panel.border = element_blank()) +
  labs(x = NULL,
       y = NULL,
       title = "Year on year change in export values, EU/Ireland 2021 (exc NR)")

valdiffs %>% 
    filter(date >= "2021-01-01", flow == "Exports", euirl != "RoW") %>% 
    select(!c(value, diff)) %>% 
    mutate(valuediff = money(valuediff)) %>% 
    pivot_wider(names_from = date, values_from = valuediff) %>% 
    arrange(divdesc, euirl) %>%   
    reactable(filterable = TRUE, compact = TRUE)


```

## Non response estimates for values

The charts and tables below detail the contribution of non response to overall trade values. Whilst the contribution is generally low, non response estimates for imports have increased markedly since January 2021.

```{r nonresponse}

nrdiv <- divs %>% 
     left_join(nonresponse) %>% 
      pivot_longer(cols = c(value,valuenr))

# Non repsonse time series line chart
nonresponse %>% 
  group_by(flow, date) %>% 
  summarise(valuenr = sum(valuenr)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = valuenr)) +
  geom_line(aes(colour = flow)) +
  scale_y_continuous(labels = label_comma(scale = 0.000001, prefix = "£", suffix = "m")) +
  scale_colour_brewer(palette = chart_palette) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = c(0.2,0.9),
        panel.grid.minor = element_blank()) +
  labs(x = NULL,
       y = "value",
       title = "Non response estimated values")

# Non response as proportion of total value bar chart
nrdiv %>% 
  # filter(flow == "Exports") %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_bar(aes(fill = name), stat = "identity") +
  facet_grid(row = ~flow) +
  scale_y_continuous(labels = comma) +
  scale_fill_discrete(labels = c("reported data", "non response estimate")) +
  theme_minimal() + 
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        panel.grid.minor = element_blank()) +
  labs(x = NULL,
       y = "value",
       title = "Non response as a proportion of all food and drink")

pcnr <- divs %>% 
  filter(date >= "2020-12-01",
         flow == "Imports") %>% 
  left_join(nonresponse) %>% 
    left_join(divcodes) %>% 
  group_by(divdesc, flow, date) %>% 
  summarise(value = sum(value),
            valuenr = sum(valuenr)) %>% 
  mutate(pcnr = percent(valuenr/(value + valuenr), accuracy = 1)) %>% 
  select(!c(value, valuenr)) %>% 
    pivot_wider(names_from = date,
                values_from = pcnr) %>% 
  left_join(divcodes) %>% 
  arrange(divcode)

totnr <- divs %>% 
    filter(date >= "2020-12-01",
         flow == "Imports") %>% 
  left_join(nonresponse) %>% 
    group_by(flow, date) %>% 
  summarise(value = sum(value),
            valuenr = sum(valuenr)) %>% 
  mutate(pcnr = percent(valuenr/(value + valuenr), accuracy = 1)) %>% 
  select(!c(value, valuenr)) %>% 
    pivot_wider(names_from = date,
                values_from = pcnr)




nonresponse %>% 
  left_join(divcodes) %>% 
  filter(date == "2021-03-01") %>% 
  mutate(valuenr = money(valuenr)) %>% 
  reactable(filterable = TRUE, compact = TRUE)

```

**Proportion of EU import value estimates attributed to non-response**

`r knitr::kable(totnr)`

`r knitr::kable(pcnr)`

## Trade by volume

### Summary

```{r summary_vol_datasets}


divsvol <- ffd %>% 
  group_by(flow, divcode, date) %>% 
  summarise(tonnes = sum(tonnes)) %>% 
  filter(date >= "2018-01-01") %>% 
  ungroup() %>% 
  left_join(divcodes) %>% 
  group_by(flow, divdesc, date) %>% 
  summarise(tonnes = sum(tonnes)) %>% 
  ungroup()


divsvol %>% 
  filter(flow == "Imports") %>%
  ggplot(aes(x = date, y = tonnes)) +
  geom_line(aes(colour = flow)) +
  facet_grid(rows = vars(divdesc), 
             cols = vars(flow),
             scales = "free",
             labeller = label_wrap_gen(width = 30)) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        strip.text.y = element_text(angle = 0),
        strip.text.x = element_blank()) +
  labs(x = NULL,
       y = "tonnes",
       title = "Import volumes by SITC2")

divsvol %>% 
  filter(flow == "Exports") %>% 
  ggplot(aes(x = date, y = tonnes)) +
  geom_line(aes(colour = flow)) +
  facet_grid(rows = vars(divdesc), cols = vars(flow),
             scales = "free",
             labeller = label_wrap_gen(width = 30)) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        strip.text.y = element_text(angle = 0),
        strip.text.x = element_blank()) +
  labs(x = NULL,
       y = "tonnes",
       title = "Export volumes by SITC2")


```

The following detailed charts include an assessment of a threshold for anomalous values in the volume data. The methodology is described at the end of this document. Essentially the thresholds are ranges which describe unusually high or low month on month changes in the volume of trade, based on historic movements in the time series. This is the approach used to derive 'escalation ranges' for risk ratings in recent emergency response dashboards.

```{r }

# # Summarise by div
# ffd <- ffd %>%
#         group_by(year, month, flow, divdesc) %>%
#         summarise(value = sum(value),
#                   tonnes = sum(tonnes)) %>%
#         rename(product = divdesc)


# Value data setup -------------------------------------------------------------
# Create thresholds
trade_val <- ffd %>% 
  mutate(date = as.Date(paste(year, month, "1", sep = "-"), format = "%Y-%m-%d")) %>% 
  group_by(product, flow) %>% 
  mutate(diff = value - lag(value),
         limupper = quantile(diff, 0.75, na.rm = TRUE) + (quantile(diff, 0.75, na.rm = TRUE) - quantile(diff, 0.25, na.rm = TRUE)) * 1.5,
         limupperub = limupper * 1.05,
         limupperlb = limupper * 0.95,
         limlower = quantile(diff, 0.25, na.rm = TRUE) - (quantile(diff, 0.75, na.rm = TRUE) - quantile(diff, 0.25, na.rm = TRUE)) * 1.5,
         limlowerub = limlower * 1.05,
         limlowerlb = limlower * 0.95)

# Nested for charting
trade_nest_val <- trade_val %>% 
                      ungroup() %>% 
                      group_by(flow, product) %>%
                      nest()

# Volume data setup ------------------------------------------------------------
trade_vol <- ffd %>% 
  mutate(date = as.Date(paste(year, month, "1", sep = "-"), format = "%Y-%m-%d")) %>% 
  group_by(product, flow) %>% 
  mutate(diff = tonnes - lag(tonnes),
         limupper = quantile(diff, 0.75, na.rm = TRUE) + (quantile(diff, 0.75, na.rm = TRUE) - quantile(diff, 0.25, na.rm = TRUE)) * 1.5,
         limupperub = limupper * 1.05,
         limupperlb = limupper * 0.95,
         limlower = quantile(diff, 0.25, na.rm = TRUE) - (quantile(diff, 0.75, na.rm = TRUE) - quantile(diff, 0.25, na.rm = TRUE)) * 1.5,
         limlowerub = limlower * 1.05,
         limlowerlb = limlower * 0.95)

# Nested for charting
trade_nest_vol <- trade_vol %>% 
                      ungroup() %>% 
                      group_by(flow, product) %>%
                      nest()


# every FFD code
selected <- unique(ffd$product)
# selected <- c("Beer", "Cheese")


```

```{r chart_functions}

threshold_plot <- function(data){
  data %>% 
    ggplot() +
    geom_line(aes(x = date, y = diff), colour = "blue", size = 0.5) +
    geom_point(aes(x = date, y = diff), colour = "blue", size = 0.5) +
    geom_line(aes(x = date, y = limupper), colour = "cyan4", size = 3, alpha = 0.25) +
    geom_line(aes(x = date, y = limlower), colour = "cyan4", size = 3, alpha = 0.25) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    scale_y_continuous(labels = comma) +
    labs(x = NULL,
         y = NULL,
         subtitle = "Thresholds") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
    # facet_wrap(~ product, scales = "free")
  
}



value_plot <- function(data){
  data %>% 
    ggplot() +
    geom_line(aes(x = date, y = value), colour = "blue", size = 0.5) +
    geom_point(aes(x = date, y = value), colour = "blue", size = 0.5) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    scale_y_continuous(labels = comma) +
    labs(x = NULL,
         y = "£000s",
         subtitle = "Trade value") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
    # facet_wrap(~ product, scales = "free")
  
}

volume_plot <- function(data){
  data %>% 
    ggplot() +
    geom_line(aes(x = date, y = tonnes), colour = "blue", size = 0.5) +
    geom_point(aes(x = date, y = tonnes), colour = "blue", size = 0.5) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    scale_y_continuous(labels = comma) +
    labs(x = NULL,
         y = "tonnes",
         subtitle = "Trade volume") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
    # facet_wrap(~ product, scales = "free")
  
}


plot_list <- function(cht1, cht2, title, subtitle){
  (cht1 / cht2) +
    plot_annotation(title = title,
                    subtitle = subtitle)
  
}

```

### Products {.tabset .tabset-pills}

#### Imports

```{r import_volume}

# diff <- trade_nest_vol %>% 
#         filter(flow == "Imports") %>%
#         filter(product %in% selected) %>%
#         mutate(tchart = pmap(list(data = data), threshold_plot),
#                vchart = pmap(list(data = data), volume_plot),
#                charts = pmap(list(cht1 = vchart, cht2 = tchart, title = product, subtitle = flow), plot_list)) 
# 
# diff$charts

```

#### Exports

```{r export_volume}

# diff <- trade_nest_vol %>% 
#         filter(flow == "Exports") %>% 
#         filter(product %in% selected) %>%
#         mutate(tchart = pmap(list(data = data), threshold_plot),
#                vchart = pmap(list(data = data), volume_plot),
#                charts = pmap(list(cht1 = vchart, cht2 = tchart, title = product, subtitle = flow), plot_list)) 
# 
# diff$charts

```

## Annex: Threshold Methodology

For each series the method for deriving an 'escalation range' was as follows:

1.  Convert data into a time series
2.  Determine how many first differences[^1] and seasonal differences[^2] need to be taken to make the time series stationary (constant statistical properties over time) using the ndiffs() and nsdiffs() functions from the forecast package in R
3.  Carry out the required differencing using the diff() function from the forecast package in R
4.  Using the data points in this stationary time series, find the threshold for outliers
5.  Look at +/-5% of this threshold to get an escalation range which allows for some uncertainty
6.  Sense check the escalation range by considering which past events would have warranted escalation given the derived range

[^1]: Taking the first difference of a time series means converting it to a time series of the differences between adjacent observations in the original time series.

[^2]: Taking the seasonal difference of a time series means converting it to a time series of the changes from one season to the next in the original time series.

The main assumptions behind this method are:

-   Past indicator data is useful to informing current escalation decisions
-   The differences of a time series can be used to detect crises related to the original time series. Specifically jumps, kinks, spikes etc... are not distinguished between in difference time series therefore the assumption is that all of these events should be identified by the escalation range.
